---
title: "data_merging/cleaning"
output: html_document
date: "2025-03-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(fuzzyjoin)
library(tidyverse)
library(plotly)
library(janitor)
library(corrplot)
```


read files
```{r}
player_play <- read.csv('data/player_play.csv')
players <- read.csv('data/players.csv')
plays <- read.csv('data/plays.csv')
combine <- read.csv('data/filtered_combine.csv')
```

get all route runners from weeks 1-9
```{r}
player_play |> 
  filter(
    !is.na(wasRunningRoute),
    wasRunningRoute == 1
  ) |> 
  dplyr::select(nflId) |> 
  group_by(nflId) |> 
  summarize(n_routes = n()) |> 
  left_join(
    y = players,
    by = "nflId"
  ) -> route_runners

route_runners <- route_runners %>%
  separate(height, into = c("feet", "inches"), sep = "-", convert = TRUE) %>%
  mutate(total_inches = feet * 12 + inches) %>%
  select(-feet, -inches)
```



all route runners + combine stats 
```{r}
route_runners_cobine <- stringdist_left_join(
  x = route_runners,
  y = combine,
  by = c('displayName' = 'Player'),
  distance_col = 'string_dist'
) %>% select(nflId,n_routes,position,displayName, weight, X40yd, X3Cone, Shuttle, total_inches)

```


looking at how missing values correlate with players performance (measured by number of routes ran)
```{r}

# all players with missing values
# 301
missing_value_players <- route_runners_cobine %>% filter(if_any(everything(), is.na))

# players with no missing values
# 192
full_data_players <- route_runners_cobine %>% drop_na()

# players with 2 or less missing values
# 313
two_missing_max <- route_runners_cobine %>% filter(rowSums(is.na(.)) <= 2)


df1 <- missing_value_players %>% select(n_routes) %>% mutate(cat = "missing")
df2 <- full_data_players %>% select(n_routes) %>% mutate(cat = "full")
df3 <- two_missing_max %>% select(n_routes) %>% mutate(cat = "2 missing")


df_combined <- bind_rows(df1, df2,df3)

# Create the violin plot
ggplot(df_combined, aes(x = cat, y = n_routes, fill = cat)) +
  geom_violin(trim = FALSE) +  # Use trim=FALSE to show the full distribution
  scale_y_continuous(expand = c(0, 0)) +  # Ensures the same scale for y-axis
  theme_minimal() +
  labs(title = "Comparison of Two Variables", y = "Value", x = "Category") +
  theme(legend.position = "none")
```
we see that players that are missing 2 values or less have a very similar distribution of number of routes ran as players that have no missing values


```{r}
cor_matrix <- cor(route_runners_cobine %>% select(where(is.numeric)), use = "complete.obs")
corrplot(cor_matrix, method = "circle")
```

imputing missing data, if we used all data, it would likely be too sparse, so we can lose a few players (hopefully uncorrelated) so we can have better data (754 missing values to 220 missing values), so we are only using players that are only missing 2 or less values
```{r}
install.packages('mice')
library(mice)

colSums(is.na(route_runners_cobine))
colSums(is.na(two_missing_max))

sum(is.na(route_runners_cobine))
sum(is.na(two_missing_max))

imputed_df <- mice(two_missing_max, method = "pmm")
imputed_rr_combine <- complete(imputed_df)

```


```{r}

```